SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-04-10 2017-Xudong-SMT/ActaInformatica.tex
a few typos corrected

\institute{Xudong Qin(\email{marsxd@gmail.com}) \and
  Zechen Hou (\email{zechen@sei.ecnu.edu.cn}) \and
  Yuxin Deng (\email{yxdeng@sei.ecnu.edu.cn}) \and
  Min Zhang (\email{mzhang@sei.ecnu.edu.cn}) \at
  Shanghai Key Laboratory of Trustworthy Computing, ECNU, China%% \\
  %% \email{marsxd@gmail.com, zechen@sei.ecnu.edu.cn, yxdeng@sei.ecnu.edu.cn, mzhang@sei.ecnu.edu.cn}
  \and
  Simon Bliudze \at
  INRIA Lille -- Nord Europe,
  40 avenue Halley,
  59650 Villeneuve d'Ascq,
  France\\
  \email{simon.bliudze@inria.fr}
  \qquad
  \orcid{0000-0002-7900-5271}
  \and
  Eric Madelaine \at
  Universit\'e C\^ote d'Azur, Inria, CNRS, I3S, 06902 Sophia
  Antipolis, France,\\
  \email{eric.madelaine@inria.fr}
  \qquad
  \orcid{0000-0002-5552-5993}
}

@@

In Algorithm~\ref{alg1}, functions \textit{filterUnsatTransitions} (in line 11) 
\textit{RefineReachableSubAutomaton} (in line 13) make use of the Z3 engine to implement the
satisfiability checking, for all the OTs computed that far.
For a transition
\[		\inferrule*[myfraction=\reddottedrule]
		{%\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
			\{\xrightarrow{\beta_j}_j\}^{j\in J}, \Pred\,, \Post}
		{s \xrightarrow {\alpha}s'}
                \]
                we need to build a predicate encoding both the OT predicate and
the information available on the possible assignments of the initial
state $s$. This is expressed as
\[Pred_{l}\wedge \bigwedge_{v\in vars(s)}\{ \bigvee_{Assign\in
  assigns(v,s)} Assign\}
\]
where each $Assign$ represents one of the possible (symbolic) value of
 the state variable $v$.

@@

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
\State $\mathit{L2}$=filterUnsatTransitions($\mathit{L,Assignments}$);
\State $\mathit{OA_2^0}$=makeReachableSubAutomaton($\mathit{InitState}(P),L2)$;
\Repeat $\mathit{OA_2^n}$=RefineReachableSubAutomaton($\mathit{OA_2^{n-1}}$)
\Until  $\mathit{OA_2^n}=\mathit{OA_2^{n-1}}$;
\State \Return $\mathit{OA_2^n}$;
%\EndIf
\end{algorithmic}  
\end{algorithm}

@@

We already have $Aut_2^0\ \subseteq Aut_1^0$.
We will proceed in two steps:
first prove by induction that $Hyp^k = (L_2^k\ \subseteq L_1^k)$ is
preserved by each application of $Refine$ (meaning that the refine function
in Algorithm~\ref{alg1} cannot go ``faster'' than in Algorithm~\ref{alg3}), then prove
that when the fixed points are reached, they are the same.

Suppose $Hyp^{k-1}$ holds and let us prove the validity of $Hyp^k$.
Consider an OT $l\ \in L_2^k$, but $l\ \notin L_1^k$.
We have \[ l\ \in L_2^k \implies l\ \in L_2^{k-1} \implies l\ \in
L_1^{k-1}\] so the transtion $l$ has been removed from $L_1^{k-1}$ during the last
iteration $Refine_1^k$. This can be caused by:
\begin{itemize}
\item
  either $Pred(l,L_1^{k-1})$ is UNSAT, because of some assignment
    $u \in Assigns(L_1^{k-1})$. But $l \in L_2^k$, so its predicate
    $Pred(l,L_2^{k-1})$ is satisfiable; this is a contradiction,
    because $Assigns(L_2^k) \subseteq Assigns(L_1^k)$ (having
    more possible assignments cannot reduce satisfiability).
  \item
    or there is a sequence
    $s_0\xrightarrow{l_1}s_1\xrightarrow{l_2}...s_{n}\xrightarrow{l}s_{n+1}$,
    with $s_0 \in States(L_1^{k-1})$, $Pred(l_1,L_1^{k-1})$ is UNSAT, and
    all further states unreachable in $L_1^{k-1}$. But as in the
    previous case, we would have $l_1$ in $L_2^k$ and not in $L_1^k$,
    reaching a similar contradiction.
\end{itemize}

@@

Clearly all these $\{l_i\}_{i \in [1..n]}$ were also satisfiable and reachable in $L_2^0$ during the ``smart'' traversal, which did not use assignments. Now consider the chain of assignment sets $\{Assigns(L_2^i)\}_{i \in [0..k]}$, used during the iteration of $Refine$ by algorithm \ref{alg3}; we argue that the specific instanciation of variables that makes $Pred(l,L_1^*)$ satisfiable was contained in $Assigns(L_2^0)$, and that if it were removed during one of the $Refine$ step in algorithm \ref{alg1}, then it would be remove similarily during algorithm \ref{alg3}. So we get a contradiction.

>>>>

\institute{Xudong Qin(\email{marsxd@gmail.com}) \and
  Zechen Hou (\email{zechen@sei.ecnu.edu.cn}) \and
  Yuxin Deng (\email{yxdeng@sei.ecnu.edu.cn}) \and
  Min Zhang (\email{mzhang@sei.ecnu.edu.cn}) \at
  Shanghai Key Laboratory of Trustworthy Computing, East China Normal University, China%% \\
  %% \email{marsxd@gmail.com, zechen@sei.ecnu.edu.cn, yxdeng@sei.ecnu.edu.cn, mzhang@sei.ecnu.edu.cn}
  \and
  Simon Bliudze \at
  INRIA Lille -- Nord Europe,
  40 avenue Halley,
  59650 Villeneuve d'Ascq,
  France\\
  \email{simon.bliudze@inria.fr}
  \qquad
  \orcid{0000-0002-7900-5271}
  \and
  Eric Madelaine \at
  Universit\'e C\^ote d'Azur, Inria, CNRS, I3S, 06902 Sophia
  Antipolis, France,\\
  \email{eric.madelaine@inria.fr}
  \qquad
  \orcid{0000-0002-5552-5993}
}

@@

In Algorithm~\ref{alg1}, functions \textit{filterUnsatTransitions} (in line 11) 
\textit{RefineReachableSubAutomaton} (in line 13) make use of the Z3 engine to implement the
satisfiability checking, for all the OTs computed that far.
For a transition
\[		\inferrule*[myfraction=\reddottedrule]
		{%\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
			\{\xrightarrow{\beta_j}_j\}^{j\in J}, \Pred\,, \Post}
		{s \xrightarrow {\alpha}s'}
                \]
                we need to build a predicate encoding both the OT predicate and
the information available on the possible assignments of the initial
state $s$. This is expressed as
\[Pred_{l}\wedge \bigwedge_{v\in vars(s)}\{ \bigvee_{Assign\in
  assigns(v,s)} Assign\}
\]
where each $Assign$ represents one of the possible (symbolic) values of
 the state variable $v$.

@@

\begin{algorithm}[h]
  \caption{``Smart'' Reachable Open Automaton Generation}
  \label{alg3}
\begin{algorithmic}[1]
\Require A pNet $P$ (cannot be a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
        Check satisfiability of $\mathit{OT}$ using the SMT solver;
        \IfThen {$SAT(\mathit{OT})$}
                {\State \{Add $\mathit{OT}$ to $L$;
                  \State Let $S'$ be the target global state of $\mathit{OT}$
	          \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
                  \}}
	\EndFor
\EndWhile
\State $\mathit{L2}$=\textit{filterUnsatTransitions}($\mathit{L,Assignments}$);
\State $\mathit{OA_2^0}$=\textit{makeReachableSubAutomaton}($\mathit{InitState}(P),L2)$;
\Repeat $\mathit{OA_2^n}$=\textit{RefineReachableSubAutomaton}($\mathit{OA_2^{n-1}}$)
\Until  $\mathit{OA_2^n}=\mathit{OA_2^{n-1}}$;
\State \Return $\mathit{OA_2^n}$;
%\EndIf
\end{algorithmic}  
\end{algorithm}

@@

We already have $Aut_2^0\ \subseteq Aut_1^0$.
We will proceed in two steps:
first prove by induction that $Hyp^k = (L_2^k\ \subseteq L_1^k)$ is
preserved by each application of \textit{Refine} (meaning that the refine function
in Algorithm~\ref{alg1} cannot go ``faster'' than in Algorithm~\ref{alg3}), then prove
that when the fixed points are reached, they are the same.

Suppose $Hyp^{k-1}$ holds and let us prove the validity of $Hyp^k$.
Consider an OT $l\ \in L_2^k$, but $l\ \notin L_1^k$.
We have \[ l\ \in L_2^k \implies l\ \in L_2^{k-1} \implies l\ \in
L_1^{k-1}\] so the transtion $l$ has been removed from $L_1^{k-1}$ during the last
iteration $Refine_1^k$. This can be caused by:
\begin{itemize}
\item
  either $Pred(l,L_1^{k-1})$ is UNSAT, because of some assignment
    $u \in Assigns(L_1^{k-1})$. But $l \in L_2^k$, so its predicate
    $Pred(l,L_2^{k-1})$ is satisfiable; this is a contradiction,
    because $Assigns(L_2^k) \subseteq Assigns(L_1^k)$ (having
    more possible assignments cannot reduce satisfiability).
  \item
    or there is a sequence
    $s_0\xrightarrow{l_1}s_1\xrightarrow{l_2}...s_{n}\xrightarrow{l}s_{n+1}$,
    with $s_0 \in States(L_1^{k-1})$, $Pred(l_1,L_1^{k-1})$ is UNSAT, and
    all further states are unreachable in $L_1^{k-1}$. But as in the
    previous case, we would have $l_1$ in $L_2^k$ and not in $L_1^k$,
    reaching a similar contradiction.
\end{itemize}

@@

Clearly all these $\{l_i\}_{i \in [1..n]}$ were also satisfiable and reachable in $L_2^0$ during the ``smart'' traversal, which did not use assignments. Now consider the chain of assignment sets $\{Assigns(L_2^i)\}_{i \in [0..k]}$, used during the iteration of $Refine$ by Algorithm \ref{alg3}; we argue that the specific instantiation of variables that makes $Pred(l,L_1^*)$ satisfiable was contained in $Assigns(L_2^0)$, and that if it were removed during one of the $Refine$ step in Algorithm \ref{alg1}, then it would be removed similarly in Algorithm \ref{alg3}. So we get a contradiction.
